---
title: "Network Psychometrics - Part 1"
format: 
  revealjs:
    theme: [default, psych-lecture-theme.scss] 
    css: psych-lecture-style.css
---

## Overview 

```{r}
#| echo: false
library(dplyr)
library(igraph)
library(qgraph)
library(tibble)
library(bootnet)
```

* Introduction to the network approach to psychometrics
* Introduction to network analysis
* Introduction to network (graphical) models

## Modern Psychometric Methods

* Latent Variable Models (IRT, CFA, EFA, etc.)
  - Reflective models
  - Formative models

## Challenges with Latent Variable Models

* Latent variable models do a poor job of specifying: 
  - The temporal precedence of the latent variable
  - The causal response processes
  - Causal associations among observed variables

## The Network Approach 

* Discontent with current methods led a group of reserachers to introduce and build on: 
  - Network theories of psychological phenomena 
  - Network (graphical) models

## Network Theories: Constructs as Dynamical Systems

* Reframe theories of psychological phenomena as a dynamic system of interactions among observable variables, which are referred to as **autonomous causal entities**.

* Most network psychometric theories pull from the literature on dynamic systems and complex systems. An example is considering depression not as a construct, but as an attractor state of the depression system. 

## Network (Graphical) Models

* Network psychometric analysis relies **heavily** on a class of statistical models called graphical models to estimate the network that underlies the observed variables. 

* One of the most common ways to estimate the underlying network is to use a model called the **Gaussian Graphical Model**, which is just a partial correlation matrix.

# Network Analysis 

## What is a Network? 

* Networks (graphs) are mathematical objects that represent connections among different entities. 

* Mathematically, networks consist of a: 
  - Set of nodes or vertices 
  - Set of edges or links

## Examples of Networks 

* Social Networks 
* Organizational Networks
* Biological Networks
* Information Networks

## What is a Psychometric Network?

* A psychometric network is a network in which the nodes are the observed variables and the edges are the conditional relationships among those observed variables.

## Network Types 

* We can broadly categorize networks by the directionality (or lack thereof) and weights of the edges:
  - Undirected vs Directed Networks
  - Unweighted vs Weighted Networks

* Psychometric network analysis typically deals with undirected and weighted networks.

## Matrix Representation of a Network

```{r}
w <- matrix(
  c(
    0, .50, .40, 0, 0, -.10,
    .50, 0, .30, -.30, 0, 0,
    .40, .30, 0, 0, 0, 0,
    0, -.30, 0, 0, .40, .30, 
    0, 0, 0, .40, 0, .20, 
    -.10, 0, 0, .30, .20, 0
  ),
  nrow = 6, ncol = 6, byrow = TRUE
)

a <- matrix(as.numeric(w != 0), nrow = 6, ncol = 6, byrow = T)
```

* The easiest way to mathematically manipulate a network is to turn it into a matrix. This matrix is referred to as an **adjacency matrix**. 

* Non-zero elements in the matrix represent connections between two nodes in the network. 

## Unweighted, Undirected Network

```{r}
qgraph::qgraph(a, color = c(bg = "#FEFBF6"), bg = "#FEFBF6")
```


## Adjacency Matrix

```{r}
a
```

## Weighted, Unirected Network 

```{r}
qgraph::qgraph(w, edge.labels = T, color = c(bg = "#FEFBF6"), bg = "#FEFBF6")
```

## Weight Matrix 

```{r}
w
```

## Analyzing Networks

* Local network properties: Metrics focused on specific parts of a network
* Global network properties: Metrics focused on the network as a whole

## Local Network Properties: Centrality 

* Centrality is a class of network metrics that measure the importance of a given node by counting and, at times, weighting the number of direct or indirect connections a node has with other nodes.

* Centrality metrics can be subdivided into: 
  - Direct connection centrality metrics
  - Indirect connection centrality metrics 

## Direct Connection Centrality Metrics

:::: {.columns}

::: {.column width="50%"}
**Degree Centrality**: 

$$\sum_{j=1}^{n}{a_{ij}}$$

**Strength**: 

$$\sum_{j=1}^{n}{\lvert w_{ij}\rvert}$$

:::

::: {.column width="50%"}
**Weighted Degree Centrality**: 

$$\sum_{j=1}^{n}{a_{ij}^{1-\alpha}\lvert w_{ij}\rvert^{\alpha}}$$

**Expected Influence**: 

$$\sum_{j=1}^{n}{a_{ij}w_{ij}}$$

:::

::::

## Direct Connection Centrality Example 

```{r}
#| echo: true
#| fig-align: center
rowSums(a) # Degree
rowSums(a*abs(w)) # Strength
rowSums(a*w) # Expected Influence
```

```{r}
#| fig-align: center
qgraph::qgraph(w, edge.labels = T, color = c(bg = "#FEFBF6"), bg = "#FEFBF6")
```

## Direct Connection Centrality `qgraph` Example 

```{r}
#| echo: true
qgraph::centralityPlot(w, include = c("Strength", "ExpectedInfluence"))
```

## Path Metrics

* Before we can calculate the indirect connection centrality metrics, we first need to understand path metrics:

:::: {.columns}

::: {.column width="50%"}

**Length**: The distance between two nodes.

$$\text{Length}(i, j)=\frac{1}{\rvert w_{ij} \lvert^{\alpha}}$$

:::

::: {.column width="50%"}

**Distance**: The shortest path between two nodes.

$$\text{Distance}(i, j)=min(w_{ik}+ \dots + w_{lj})$$

:::

::::

## Indirect Connection Centrality Metrics 

:::: {.columns}

::: {.column width="50%"}

**Closeness**: 

$$\frac{1}{\sum_{j=1}^{n}\text{Distance}(i,j)}$$

**Betweeness**:

$$\sum_{<j,k>}\frac{\text{# shortest paths from j to k that go through i}}{\text{# shortest paths between j and k}}$$

:::

::: {.column width="50%"}


**Expected Influence Two**: 

$$\sum_{j=1}^{n}a_{ij}w_{ij}+\sum_{j=1}^{n}a_{ij}w_{ij}\sum_{k=1}^{n}a_{jk}w_{jk}$$

:::

::::

## Indirect Connection Centrality Example 

```{r}
#| echo: true
#| fig-align: center
(rowSums(w) + a%*%rowSums(w)) |> t()  # Expected Influence 2
qgraph::centralityPlot(w, include = c("all"))
```

## Local Network Properties: Clustering

* Clustering is a metric that looks at how often two nodes that are connected to the same focal node are also connected to one another: the friend of my friend is also my friend.

* I haven't seen this used a whole lot in psychometric network analysis. 

## Global Network Properties: Density

* It is often of interest to undestand how connected your network is. This can be calculated using a network metric called **density**.

$$\text{Density}=\frac{\text{# Observed Edges}}{\text{# Possible Edges}}=\frac{\sum_{<i, j>}a_{ij}}{n(n-1)}$$

## Global Network Properties: Small Worlds

* Six degrees of Kevin Bacon
* Small world network models are characterized by high clustering and short average path lengths--this does not happen when networks form completely at random.  

## Calculating Density and Small World Index in R

```{r}
#| echo: true
toy_graph <- qgraph::qgraph(w, DoNotPlot = TRUE) # Create your graph
toy_graph |> igraph::as.igraph() |> igraph::edge_density() # Calculate density
qgraph::smallworldIndex(toy_graph)$index # Calculate small world index
```

## Global Network Properties: Community Detection

* In network analysis, community detection is a set of alogrithms that attempt to find hidden communities within a network by identifying sets of nodes that have more connections among themselves than they have with other nodes outside of the set. 

* Common algorithms: 
  - Walktrap Algorithm
  - Louvain Algorithm
  - Fast-Greedy Algorithm

## Community Detection Example

```{r}
#| echo: true
# Need to take the absolute value of the weighted network
toy_graph <- qgraph(abs(w), DoNotPlot = TRUE)
toy_graph |> igraph::as.igraph() |> igraph::walktrap.community()
```

# Network Models

## What makes Network Psychometrics Pscyhometric? 

* Psychometric network analysis is "the analysis of multivariate psychometric data using network structure estimation and network description" (Borsboom et al., 2021).

* Estimating the network structure from psychometric data is what makes psychometric network analysis psychometric.

## Estimating Psychometric Networks

* Rather than assume that our data is generated by a some latent variable, we assume that our data was generated by some unobserved system (network) of interactions among our variables.

* Psychometric network analysis uses a network model called a **Pairwise Markov Random Field (PMRF) Model** to estimate the underlying network. 

## Pairwise Markov Random Fields 

* PMRF models are network models that use the conditional relationships among one's variables (e.g. the relationship between x and y controlling for z) to estimate the edges between two nodes (variables).

* Types of PMRF models: 
  - Guassian Graphical Model (GGM)
  - Ising Model
  - Mixed Graphical Models

## Gaussian Graphical Model and Partial Correlations

* Given a set of variables, the GGM uses the partial correlations among those variables as a means to estimate the edges. 

* Partial correlation for three variables: x, y, & z:

$$r_{xy.z}=\frac{r_{xy}-r_{xz}r_{yz}}{\sqrt{1-r_{xz}^2}\sqrt{1-r_{yz}^2}}$$

## Estimating the Saturated GGM 

* To estimate a saturated GGM (all correlations among all variables), we just need the covariance matrix:

$$\mathbf{K}=\mathbf{\Sigma^{-1}}$$

$$\omega_{ij}=-\frac{k_{ij}}{\sqrt{k_{ii}}\sqrt{k_{jj}}}$$

## Example of the Saturated GGM {.scrollable}

```{r}
sigma_mat <- matrix(c(1, .50, .25, .50, 1, .50, .25, .50, 1), nrow = 3, ncol = 3, byrow = T)
k <- solve(sigma_mat)
scale_mat <- sqrt(diag(k))^-1 |> diag(3)
omega <- scale_mat%*%-k%*%scale_mat + diag(3)
```

:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: true
# Correlation Matrix
sigma_mat

# Partial Correlation Matrix
round(omega, 2)
```

:::
::: {.column width="50%"}

```{r}
qgraph::qgraph(omega, color = c(bg = "#FEFBF6"), bg = "#FEFBF6")
```

:::
::::

## Model Selection for GGM

* Data are noisy and it is likely there will be partial correlations by chance. We want to weed those correlations out using some kind of model selection.

* Several ways to select the "best" GGM for your data: 
  - Thresholding & Pruning
  - Model Search using information criteria
  - Regularization using the GLASSO

## Some Critiques of Psychometric Network Analysis

* **Model Selection**: Are network models appropriate for psychological data?
* **Study Design**: Can we infer anything about a dynamical system from cross-sectional data? 
* **Estimation Reliability**: How stable/replicable are psychometric networks?  
* **Interpretation of Measures**: Can we even interpet network metrics applied to psychometric networks?

::: aside
See Neal et al. (2022). Critiques of network analysis of multivariate data in psychological science. 
:::



